Learning Auditory Similarities Between Instrumental Playing Techniques

### 2019-10-05 paper/paper.tex
cite faure

Despite their widespread use, linguistic references to musical timbre suffer from limited adequacy.
On the one hand, adjectives such as \emph{bright} or \emph{rough} are prone to misunderstanding, as they do not prescribe any musical gesture that is capable of achieving them.
On the other hand, the sole mention of playing technique does not determine any intended effect; and therefore, fails to generalize to the entire instrumentarium.
For instance, although the term \emph{breathy} alludes to a playing technique that is peculiar to wind instruments, a cellist may accomplish a seemingly breathy timbre by bowing near the fingerboard, i.e., \emph{sul tasto} in the classical terminology.

@@

In this paper, we propose a machine listening system which computes the dissimilarity in timbre between any two audio samples.
This system composes two stages: unsupervised feature extraction and supervised metric learning.
The feature extraction stage is a nonlinear map which relies on the joint time--frequency scattering transform \cite{anden2019joint}, followed by per-feature Gaussianization \cite{lostanlen2018relevance}.
It aims at encoding patterns of spectrotemporal modulation in the query while offering numerical guarantees of stability to local deformations.
The metric learning stage is a linear map, optimized via large-margin nearest neighbors (LMNN), which recombines scattering coefficients so as to fit a training set of human judgments.

@@

On the other hand, the field of music psychology investigates timbre with the aim of discovering its physiological and behavioral foundations \ml{vl cite}.
In this setting, prior knowledge, however accurate, of instrumentation does not suffice to conduct a study.
Rather, each excerpt must be played back to multiple human listeners.
Yet, collecting subjective responses to acoustic stimuli is a tedious and unscalable procedure, thus restricting the size of the musical corpus under study.
Such restriction hampers the applicability of iterative representation learning algorithms, such as stochastic gradient descent in artificial neural networks.
Nevertheless, advanced methods in electrophysiology allow to observe the firing patterns of biological neurons in the presence of controlled stimuli.
This observation, originally carried out on the ferret, has led to a comprehensive mapping of the primary auditory cortex, in terms of its spectrotemporal receptive fields (STRF) \cite{depireux2001amphyssoc}.

>>>>

Despite their widespread use, linguistic references to musical timbre suffer from limited adequacy.
On the one hand, adjectives such as \emph{bright} or \emph{rough} are prone to misunderstanding, as they do not prescribe any musical gesture that is capable of achieving them \cite{faure1996icmpc}.
On the other hand, the sole mention of playing technique does not determine any intended effect; and therefore, fails to generalize to the entire instrumentarium.
For instance, although the term \emph{breathy} alludes to a playing technique that is peculiar to wind instruments, a cellist may accomplish a seemingly breathy timbre by bowing near the fingerboard, i.e., \emph{sul tasto} in the classical terminology.

@@

In this paper, we propose a machine listening system which computes the dissimilarity in timbre between any two audio samples.
This system composes two stages: unsupervised feature extraction and supervised metric learning.
The feature extraction stage is a nonlinear map which relies on the joint time--frequency scattering transform \cite{anden2015joint}, followed by per-feature Gaussianization \cite{lostanlen2018relevance}.
It aims at encoding patterns of spectrotemporal modulation in the query while offering numerical guarantees of stability to local deformations.
The metric learning stage is a linear map, optimized via large-margin nearest neighbors (LMNN), which recombines scattering coefficients so as to fit a training set of human judgments.

@@

On the other hand, the field of music psychology investigates timbre with the aim of discovering its physiological and behavioral foundations \ml{vl cite}.
In this setting, prior knowledge, however accurate, of instrumentation does not suffice to conduct a study.
Rather, each excerpt must be played back to multiple human listeners.
Yet, collecting subjective responses to acoustic stimuli is a tedious and unscalable procedure, thus restricting the size of the musical corpus under study.
Such restriction hampers the applicability of iterative representation learning algorithms, such as stochastic gradient descent in artificial neural networks.
Nevertheless, advanced methods in electrophysiology allow to observe the firing patterns of biological neurons in the presence of controlled stimuli.
This observation, originally carried out on the ferret, has led to a comprehensive mapping of the primary auditory cortex, in terms of its spectrotemporal receptive fields (STRFs) \cite{depireux2001jneur}.

Our model borrows from both aforementioned lines of research in music information retrieval and music psychology.
Indeed, the joint time--frequency scattering transform can be regarded as a deep convolutional network with fixed synaptic weights; or, alternatively, as an idealized computational model for STRFs \cite{anden2019joint}.
